---
title: " "
author: "January Weiner"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    css: "files/style.css"
---

# 

## #About

January Weiner `<january.weiner@gmail.com>`

Core Unit Bioinformatics, Berlin Institute of Health


## Course materials

https://github.com/january3/Radvanced2021

## What I want to give you

 * Make R friendly
 * Indoctrination
 * Ability to write code for others
 * Practical tips and tricks

## Course motto

> R is a shockingly dreadful language for an exceptionally useful data
> analysis environment. The more you learn about the R language, the worse
> it will feel. The development environment suffers from literally decades
> of accretion of stupid hacks from a community containing, to a
> first-order approximation, zero software engineers. R makes me want to
> kick things almost every time I use it.

Tim Smith, Kevin Ushey, [arrGh: a newcomer’s (angry) guide to R](http://arrgh.tim-smith.us/)

## How the course will look like

 * Lectures
 * Excercises
 * Scripts
 * Your own projects


# Why "good coding practices"?


## Rule #1: Language is for communication

Code is a tool of your science, science is all about communication, therefore:

 * Your code will be inspected by others
 * Documentation is not only for you
 * Coding is a part of your science, then make it so!
 * Reproducibility matters

 
## Literate programming (Donald E. Knuth)

I believe that the time is ripe for significantly better documentation of
programs, and that we can best achieve this by considering programs to be
works of literature. Hence, my title: "Literate Programming."

Let us change our traditional attitude to the construction of programs:
Instead of imagining that our main task is to instruct a computer what to
do, let us concentrate rather on explaining to human beings what we want a
computer to do.

## Literate Programming (Donald E. Knuth)

The practitioner of literate programming can be regarded as an essayist,
whose main concern is with exposition and excellence of style. Such an
author, with thesaurus in hand, chooses the names of variables carefully
and explains what each variable means. He or she strives for a program that
is comprehensible because its concepts have been introduced in an order
that is best for human understanding, using a mixture of formal and
informal methods that reinforce each other.
 

## Rule #2: Document your code

*Documentation is like pizza: if it's good, it's really, really good. If
it's bad, it's still better than nothing.* [Anonymous]

 * make documenting your code as easy as possible
 * first document the why, then the what and the how
 * make documentation easily searchable (use grep, find & co)

## Rule #3: Stick to conventions

 * adopt a naming and formatting style convention, and stick to it
 * keep the same names for variables over and over
 * keep the same project / workspace structure

## Rule #4: Keep your code clean

 * meaningful variable names
 * refactor your code
 * organize your code (packages!)
 * keep your functions short
 * keep a line length of approximately 80 chars

## Rule #5: Simplify!

 * make your code as simple as possible
 * make your functions versatile and simple
 * use simple data types if possible
 * don't overdo it!

## Rule #6: You never know

 * what your code evolves into
 * when you will need to publish it
 * when someone will want to see it

…so prepare in advance!

## Exercise Session [day 1/1](../Exercises/day_1.md)
 
# Specifics (the gory details)

## Before you start

 * think what you want to do
 * make a plan, use paper and pencil
 * use google and stackoverflow, do not reinvent the wheel
 * learn to use (r)markdown


## How to write code

 * use a naming convention (eg. with.period, with\_underscore or camelCase)
 * strive for meaningful variables…
 * …but keep them concise
 * make function and variable names easy to distinguish (e.g. noun vs verb
   or use a prefix)
 * make it easy to distinguish less important (e.g. local) variables and
   functions from those which are important
 * package your code in functions, don't use globals
 * package your functions in packages, don't use scripts 

##

*Firstly, don't call your matrix 'matrix'. Would you call your dog 'dog'?
Anyway, it might clash with the function 'matrix'.*

– Barry Rowlingson, R-help (October 2004)


## Keep your syntax clean and standard

 * there are many coding styles, but they have lots in common
 * spaces around operators
 * always use `<-` for assignment (never `=`)
 * use double hash (`##`) if comments starts in a new line 
 * avoid semicolons
 * use curly braces even when not necessary (e.g. in `if() … else …`)


## Organize your work space

 * readme.rmd or readme.txt files in every directory
 * workspaces / projects and always know where you are
 * standard project structure
 * meaningful file names
 * standard file extensions (`.R`, `.rda`, `.tsv`, `.csv`)

## A matter of taste

 * use `stop()`, `try()`/`tryCatch()`, `warn()` and `message()`
 * don't use `attach()` (use tidyverse instead!)
 * row names, should one use them?
 * use `rmarkdown` / `roxygen2` to document your functions on the go
 * use spaces to indent your code (not tabs)
 * keep track of the changes you make to your code and project

## Trying

"Try. Fail. Try again. Fail again. Fail BETTER" (Samuel Beckett)

```r
res <- try(exp(42))
res <- try(exp("forty two"))

if(inherits(res, "try-error")) {
  warning("The operation did not work")
}
```

## Documenting functions with roxygen

```r
#' Function for adding two numbers
#' @param a first number
#' @param b second number
#' @return sum of the numbers
add2nums <- function(a, b) {
  a + b
}
```

## More on that

* always start with `#' `
* `@param` – parameter
* `@return` – return value
* `@export`– the function should be accessible to users
* `@seealso` – other related functions
* `@examples` – section with examples
 
## Exercise Session #2

  "Writing and documenting functions"



# R as a functional language

## Functional languages

  * Clean: Haskell, Clean, PureScript
  * Impure: C++, Java, Python, R
  * functional programming paradigm:
      * functions are called by arguments only and do not modify any
        environmental variables
      * vectorization
      * higher order functions, closures and anonymous functions

## Pure functions

Pure functions do not modify their environment and their return value
determistically depends on the arguments. This is often not
possible (think: randomization, reading data).

However, they have many advantages and it is worth pursuing them.

## Everything is a function

In R, almost everything is a function, including most operators:

```r
"+"(1, 2)

a <- LETTERS[1:10]
"["(a, 3) # prints out "C"
```

## Defining your own operators

Use `%...%` to define your own operators:

    %M% <- function(a, b) setdiff(a, b)

    1:10 %M% 3:4 
    # same as: setdiff(1:10, 3:4)

    %P% <- function(a, b) paste0(a, b)

You can also use special characters:

    `%-%` <- function(a, b) setdiff(a, b)

## Strict vs lazy evaluation

Consider the following statements:

    print(length(1/0))

    f <- function(a=b) { b <- 17 ; print(a) }
    f()


  Which one will throw an error? Why?

## Exercise Session #3

  "R as functional language"

## Vectorization

Means operating on vectors and matrices rather than on single values.
Instead of using a `for` loop to walk over all elements of a vector, one
calls a function which operates on the vector as a whole. This is how
standard R functions (such as `+`) work.

Some vectorized functions:

 * practically everything
 * most of the basic functions like `sum`, `log`, `sd`, `mean`
 * `colMeans`, `colSums`, `rowMeans`, `rowSums` do exactly what they say
 * make sure your functions are vectorized as well!
 * if you can't find a vectorized function, use `*map*` / `*apply`

## Vectorization

Use `*apply` and, from tidyverse's `purrr` package, `*map*` instead of
loops. Loops in R are almost never necessary.

However, for some tasks, don't be a purist. `for` is there for a reason
(although you might consider the `walk` family of functions from `purrr`).

Remember, there are no single values in R. Even `1` is a vector (of length
1).

## Vectorization

```{r, eval=FALSE}
data(iris)
## per species median
tapply(iris[,1], iris$Species, median)

species <- unique(iris$Species)
names(species) <- species
mysummary <- function(sp, species_vec, dataset, col) {
  x <- dataset[[col]]
  x <- x[ sp  == species_vec ]
  c(mean(x), median(x), IQR(x))
}

lapply(species, mysummary, iris$Species, iris, "Sepal.Width")
```

## Vectorization – the tidyverse solution

```{r, eval=FALSE}
library(tidyverse)
iris %>% group_by(Species) %>% 
  summarise(median=median(Sepal.Width), mean=mean(Sepal.Width))
```

# Common pitfalls

## Floating point

Are all of the following statements true?

    0.1 * 10 * 3 / 3 == 1
    3 / 3 * 0.1 * 10  == 1
    3 * 0.1 / 3 * 10  == 1

Solution: use epsilon.

    eps <- 1e-12
    abs((3 * 0.1 / 3 * 10) - 1) < eps

## Exercise Session 3

  "Vectorization and floating point operations"

## Factors, factors everywhere

Factors are "tricky little devils". Be especially wary of factors that look
like numbers. For example:

    foo <- data.frame(ID=c(10:1, "a"))

Factors are often silenty translated either to a numeric or character type.
Strange things can happen then. Solution:

    options(stringsAsFactors=FALSE)


## data.frame with one column

R's default data frames quietly turn to a vector when only one column is
selected. This is usually not the desired outcome and can be avoided with

    foo[, 1, drop=FALSE]

## Partial matching behind `$`


```r
a <- list()
a$coefficient <- 2
a$c # result: 2
```

Solution:

    options(warnPartialMatchDollar = TRUE)
     
## The Namibia problem

By default, "NA" in data read means "not available", so if "NA" is a valid
string (e.g. abbreviation of "Namibia"), trouble ensues.

Solution:

```r
na.strings=NULL
```

## Precedence

    What does the following do?

```r
n <- 10
for(i in 1:n-1) print(i)
```

## Typos

What *exactly* will happen here?

```r
a <- FALSE

if(a = TRUE) {
  print("a is true!")
}
```

## & vs &&

* `&&`: operator taking two values 

* `&`: vectorized version

```{r}
a <- c(TRUE, TRUE)
b <- c(TRUE, FALSE)
a && b
a & b
```

## & vs &&

```{r}
if(a && b) { print("TRUE!") }
if(a & b) { print("TRUE!") }
```

## Forgetting to use `any`

```{r}
vec <- c(1, 2, NA, 3, 4)

if(is.na(vec)) { print("NA's found!") } else { print("No NA's!") }

if(any(is.na(vec))) { print("NA's found!") } else { print("No NA's!") }
```

## Changing one element may change the type of a vector

```{r}
a <- 1:10
typeof(a)
a[1] <- "a"
typeof(a)
```

## Confusing `length` with `nchar`

```{r}
a <- "January Weiner"
length(a)
nchar(a)
```

## Be paranoid

A Software Tester walks in a bar...


# Functions

## What is a function?

```{r}
myfunction <- function(arg1, arg2) {
  ret <- sum(arg1, arg2)
  return(ret)
}

myfunction(1, 2)
a <- 1
b <- 2
myfunction(a, b)
```

## Confusion about arguments

```{r}
input <- "myfile.txt"

myfunction <- function(input) {
  print(input)
}

myfunction(input)
myfunction("something else entirely")
another_input <- "another one!"
myfunction(another_input)
```

## Default arguments

```{r}
myfunction <- function(arg1=1, arg2=3) {
  ret <- sum(arg1, arg2)
  return(ret)
}

myfunction(a, b)
myfunction(a)
myfunction()
```

## "return" is not necessary

Last statement of a function is its return value

```{r}
myfunction <- function(arg1, arg2) {
  arg1 + arg2 # same as return(arg1 + arg2)
}
```

## A slightly more complex example

```{r}
myfunction <- function(x, operation="sum") {

  if(operation == "sum") {
    ret <- sum(x)
  } else if(operation == "mean") {
    ret <- mean(x)
  }

  return(ret)
}
```

## Matching arguments

```{r}
myfunction <- function(x, operation="sum") {

  operation <- match.arg(operation, c("sum", "mean"))

  if(operation == "sum") {
    ret <- sum(x)
  } else if(operation == "mean") {
    ret <- mean(x)
  }

  return(ret)
}

myfunction(1:100, "s")
```

## Ellipsis

```{r}
myfunction <- function(x, operation="sum", ...) {

  operation <- match.arg(operation, c("sum", "mean"))

  if(operation == "sum") {
    ret <- sum(x, ...)
  } else if(operation == "mean") {
    ret <- mean(x, ...)
  }

  return(ret)
}
input <- c(1:100, NA)
myfunction(input, "s")
myfunction(input, "s", na.rm=TRUE)
```

## Ellipsis

```{r}
myfunction <- function(x, ...) {
  args <- list(...)

  print(unlist(args))
}

myfunction(1:100, 1, 12)
myfunction(1:100, a=1, b=12, na.rm=TRUE)
```

## Anonymous functions

Pretty much like with regular variables, you don't have to create a
variable to use the value:

```r
a <- 2
a + 3
2 + 3 ## both are equivalent
```

Similarly:

```r
add <- function(a, b) { a + b }
add(1, 2)

(function(a, b) { a + b })(1, 2)
```


